Marlowe: implementing and analysing financial contracts\thanks{Supported by IOHK.}

### 2019-12-04 papers/marlowe2019/paper.tex
Add draft for formal verification section

We can also use proof assistants like Isabelle to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract). In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.

@@

\emph{ToDo}
Any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.

>>>>

We can also use proof assistants to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract). In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.

Currently, we have translated the Haskell Marlowe semantics to Isabelle as closely as possible. The main differences between the Isabelle and the Haskell semantics are that:
\begin{itemize}
    \item The Isabelle semantics uses \emph{integers for identifiers} and the Haskell semantics uses strings. We do this because handling integers in Isabelle is easier and we only compare them for equality and order (their value is irrelevant).
    \item We use a \emph{custom implementation of maps and sets} that use lists of tuples and lists of elements respectively. We do this because Isabelle already provides many theorems that are proven for lists, so it considerably facilitates the automation of certain proofs.
\end{itemize}

\subsubsection{Termination proof}

As part of the formalisation of the semantics, Isabelle proves termination automatically for most functions. The only function for which termination is not trivial is \texttt{reductionLoop}. The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so it requires proving that \texttt{reduceContractStep} will eventually do so.

The way we proved that this is the case was by establishing a measurment for the size of a pair of \texttt{Contract} and \texttt{State}. And we defined such measurement as follows:

\begin{verbatim}
fun evalBound :: "State => Contract => nat" where
"evalBound sta cont = length (accounts sta) + 2 * (size cont)"
\end{verbatim}

\texttt{size} is a measurement already generated automatically by Isabelle for custom data types.

The reason why we need the number of accounts in the \texttt{State} for the measurement is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}.

And the reason why we multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one. By multiplying the size of the \texttt{Contract} by two, its decrease compensates the increase in the size of the list of accounts and, thus, it ensures that the measurement will decrease even in that scenario.

\subsubsection{Valid state and positive account preservation}

Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of state but, nevertheless, they make no sense given what these lists represent. In particular:

\begin{enumerate}
    \item The lists of tuples are representing a maps, so it does not make sense for two elements of the list to have the same first element (key).
    \item Additionally, we want to minimise the number of ways in which a map can be represented as a list, so we require the first elements of the elements of the list to be sorted ascendingly. This allows us to use standard equality for comparing maps, for example.
    \item We only expect accounts to store positive amounts money, so the second element of the pairs of the list of accounts must be a positive integer. Whenever an account becomes empty we just delete the account from the list.
\end{enumerate}

The two first invariants we define as \texttt{validState}. And the latest we define as follows:

\begin{verbatim}
fun positiveMoneyInAccountOrNoAccount
             :: "AccountId => Token => Accounts => bool" where
"positiveMoneyInAccountOrNoAccount accId tok mlist =
    (case MList.lookup (accId, tok) mlist of
       None => True
     | Some x => x > 0)"
\end{verbatim}

And it must be \texttt{True} independently of the \texttt{AccountId} and \texttt{Token}.

We prove that, if the properties are true for a given input \texttt{State}, then the semantics will preserve the properties in the output \texttt{State}.

\subsubsection{Quiescent result}

There is a class of \texttt{Contract}s that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well. We call this state \textit{quiescent} and we define it as follows:

\begin{verbatim}
fun isQuiescent :: "Contract => State => bool" where
"isQuiescent Close state = (accounts state = [])" |
"isQuiescent (When _ _ _) _ = True" |
"isQuiescent _ _ = False"
\end{verbatim}

We prove that if the input \texttt{State} is valid and all accounts are positive, then the result of a transaction is always quiescent.

\subsubsection{Money preservation}

One of the dangers of using smart-contracts is that, a badly written one may lock its funds forever. We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money. By the end of the contract, all the money should be distributed back somehow to a subset of the participants.

We state the property as follows:

\begin{verbatim}
lemma computeTransaction_preserves_money :
"validAndPositive_state state
  ==> moneyInState state + moneyInInputs (inputs tra)
        = moneyInTransactionOutput state (inputs tra)
                (computeTransaction tra state contract)"
\end{verbatim}

The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.

@@

Another property of Marlowe is that any given finite \texttt{Contract} has an implicit finite bound on the maximum number of \texttt{Transaction}s that it accepts. This is a convenient property because:

\begin{itemize}
    \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker partcipant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
    \item It bounds the length of traces (the number of transactions) that symbolic execution (see Section~\ref{subsec:static}) needs to explore (more on this below).
\end{itemize}

We state the property as follows:

\begin{verbatim}
lemma playTrace_only_accepts_maxTransactionsInitialState :
"playTrace sl c l = TransactionOutput txOut
  ==> length l <= maxTransactionsInitialState c"
\end{verbatim}

\noindent
where \texttt{maxTransactionsInitialState} is essentially the maximum number of nested \texttt{When} in the contract plus one.

This property implies that any trace that is longer than this is guaranteed to produce at least one error. And because transactions that produce an error do not alter the state of the contract, such a list of transactions (trace) will be equivalent to a list of transactions that does not have the erroneous transaction.

In other words, for every list of \texttt{Transaction}s, there exists at least one list of \texttt{Transaction}s with length less or equal to \texttt{maxTransactionsInitialState} that has the same effect on the contracts.

Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.
